\chapter{Introduction}
\label{chapter:introduction}

Nowadays, computing science intertwines with various fields of studies posing new challenges for the people of the industry. Every aspect of science is dominated by the technology, just like the everyday lives across the world.

A good example of such interdisciplinary connection is a situation when the companies' managers are required to answer a difficult optimisation question. These businesses mostly are made of different branches, which require transferring of goods between them. Each element of the product could be manufactured in a various departments and is needed in the last part of the assembly in the main office. It could be easier when they produce the whole product in one location, yet seldom it is better to divide the responsibility for specialists in each sector. Although, this process generates som extra costs for the company, it would be crucial to minimise the expenses during assignment of the branches to the locations. Trying to resolve this by hand could be a long process due to the complexity of the problem. However, with usage of computers, it could be answered in a shorter time. The result may not be the optimal one in each case, but it should meet most given assumptions, which is enough to put it into real life usage.

In computer science, this dilemma can be stated as the Quadratic Assignment Problem (QAP).  The QAP is a combinatorial optimisation problem, which was defined by Koopmans and Beckmann in 1957 \cite{koopmans-beckmann1957}, and is a generalisation of an assignment problem. It is NP-hard thus even for small instances finding results is done by approximate methods.

The QAP can be formulated as follows: Given $ n $ different facilities ($F$) and $ n $ different locations ($L$), a weight function $ w: F \times F \mapsto R $ between facilities and a distance function $ d: L \times L \mapsto R $ between locations, find the assignment minimising this cost function:

\begin{equation}
min \sum_{a, b \in F} w(a, b) * d( f(a), f(b))
\end{equation}

Over the years, a number of methods for solving this problem were established. Nevertheless, there is still a place for improvement and experts are searching for new ways of resolving that. Inspired by their works this thesis tries to find an innovative method using living organisms.

The \textit{Physarum polycephalum}, also known as a true slime mould, is a plasmodial organism of yellow colour. Its single cell body is considered to be the biggest in the world \cite{stephenson1994myxomycetes}. Taking into account the current state of classification, it belongs to the Kingdom Protista, however, this used to frequently change due to the fact that it does not exactly match any of the recognised kingdoms. The \textit{Physarum polycephalum} move very slowly, in a branching pattern as it forages for new food sources. It ingests bacteria, fungal spores and during the experiments --- oatmeals.

\input{chapters/introduction/motivation.tex}

\input{chapters/introduction/goal.tex}

\input{chapters/introduction/chapters.tex}

\input{chapters/introduction/distribution.tex}
